Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex

Using this model, we may view the problem of estimating the similarity $Y$ of
items $A$ and $B$ that most likely produced ratings $R_A$ and $_B$ as
that of finding the Maximum Likelihood Estimate of $Y | \vec{R}$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|\vec{R}=\vec{r}) 
\\&= \argmax_y\left[\frac{P(\vec{R}=\vec{r} |
Y=y)P(Y=y)}{P(\vec{R}=\vec{r})}\right]
\\&= \argmax_y\left[P(\vec{R}=\vec{r}|Y=y)P(Y=y)\right]
\\&=
\argmax_y\left[\frac{1}{2\pi\sqrt{\left|\mathbf{\Sigma_y}\right|}}\exp{\left(-\frac{1}{2}
(\vec{r} - \vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r})\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu_s)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[(\vec{r} -
\vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r}) + \frac{(y -
\mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&=
\argmin_y\left[\vec{r^\prime}^\mathrm{T}\mathbf{\Sigma_y}^{-1}\vec{r^\prime}
 + \frac{(y - \mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&= \argmin_y\left[\frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right],
\end{align}
where we used the fact that
\begin{align}
\mathbf{\Sigma_y}^{-1} = \frac{1}{\sigma_r^2}
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]^{-1} =
\frac{1}{\sigma_r^2}\frac{1}{1 - y^2}\left[ \begin{array}{cc}  1 & -y \\ -y & 1
\end{array} \right].
\end{align}
and that
\begin{align}
\left|\mathbf{\Sigma}\right| = \sigma_r^2(1 - y^2)
\end{align}
along with the substitution $\vec{r^\prime} = (r_A^\prime~~r_B^\prime) = \vec{r}
- \vec{\mu_r}$. Next, we find the minimum by taking the derivative with respect
to $y$ and setting it to zero:
\begin{align}
\frac{d}{dy}\left[ \frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right] &= 0 \\
-r_A^\prime r_B^\prime y^2 + ({r_A^\prime}^2 + {r_B^\prime}^2)y - r_A^\prime
r_B^\prime + \frac{\sigma_r^2}{\sigma_s^2}(1 - y^2)^2(y - \mu_s) - \sigma_r^2(1
- y^2)y &= 0.
\end{align}
This final expression implicitly yields the MLE of $y$ as a function of the
ratings $r_A$ and $r_B$. That is, by constraining the parameters $\mu_r$,
$\sigma_r$, $\mu_s$, and $\sigma_s$ to realistic values, the equation
can be used to efficiently estimate $y$ given a pair of ratings.

@@

\section*{Conclusion}
\begin{thebibliography}{9}

>>>>

Using this model, we may view the problem of estimating the similarity $Y$ of
items $A$ and $B$ that most likely produced ratings $R_A$ and $R_B$ as
that of finding the Maximum Likelihood Estimate of $Y | \vec{R}$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|\vec{R}=\vec{r}) 
\\&= \argmax_y\left[\frac{P(\vec{R}=\vec{r} |
Y=y)P(Y=y)}{P(\vec{R}=\vec{r})}\right]
\\&= \argmax_y\left[P(\vec{R}=\vec{r}|Y=y)P(Y=y)\right]
\\&=
\argmax_y\left[\frac{1}{2\pi\sqrt{\left|\mathbf{\Sigma_y}\right|}}\exp{\left(-\frac{1}{2}
(\vec{r} - \vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r})\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu_s)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[(\vec{r} -
\vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r}) + \frac{(y -
\mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&=
\argmin_y\left[\vec{r^\prime}^\mathrm{T}\mathbf{\Sigma_y}^{-1}\vec{r^\prime}
 + \frac{(y - \mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&= \argmin_y\left[\frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right],
\end{align}
where we used the fact that
\begin{align}
\mathbf{\Sigma_y}^{-1} = \frac{1}{\sigma_r^2}
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]^{-1} =
\frac{1}{\sigma_r^2}\frac{1}{1 - y^2}\left[ \begin{array}{cc}  1 & -y \\ -y & 1
\end{array} \right].
\end{align}
and that
\begin{align}
\left|\mathbf{\Sigma}\right| = \sigma_r^2(1 - y^2)
\end{align}
along with the substitution $\vec{r^\prime} = (r_A^\prime~~r_B^\prime) = \vec{r}
- \vec{\mu_r}$. Next, we find the minimum by taking the derivative with respect
to $y$ and setting it to zero:
\begin{align}
\frac{d}{dy}\left[ \frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right] &= 0 \\
-r_A^\prime r_B^\prime y^2 + ({r_A^\prime}^2 + {r_B^\prime}^2)y - r_A^\prime
r_B^\prime + \frac{\sigma_r^2}{\sigma_s^2}(1 - y^2)^2(y - \mu_s) - \sigma_r^2(1
- y^2)y &= 0.
\end{align}
This final expression implicitly yields the MLE of $y$ as a function of the
ratings $r_A$ and $r_B$. That is, by constraining the parameters $\mu_r$,
$\sigma_r$, $\mu_s$, and $\sigma_s$ to realistic values (e.g. $\sigma_r,
\sigma_s > 0$), this expression defines a surface that can be used to
efficiently estimate $y$ given a pair of ratings\footnote{Note that this
equation neglects the constraint that $-1 \leq y \leq 1$. As a consequence, some
combinations of ratings and parameters can yield MLEs for $y$ that exceed $1$ in
magnitude. This is easily fixed, however, by applying this additional
constraint.}.

\subsection*{User Predictivity}

As already mentioned, the estimator $PearSim$ is unable to say much about the
similarity of two items when only provided ratings from a single common user. In
this case, $PearSim$ yields $\pm1$ (or the indeterminate value $0/0$ in the case
that both ratings are zero) depending on whether the ratings have the same sign
or not. The MLE provided by the above model rectifies this by generating
similarity estimates in the range $[-1, +1]$ given a single pair of ratings. We
therefore consider each user who rates a pair of items as implicitly predicting
the similarity of the items. A key discovery of the present work is that the
accuracy of users' predictions in a training set is correlated with the accuracy
in a test set. We exploit this finding by ascribing a quantity we call {\em user
predictivity} to each user defined as
\begin{align}
p_u = e^{-\epsilon_u/\lambda},
\end{align}
where $\lambda$ is a parameter to be learned and
\begin{align}
\epsilon_u = \frac{1}{\left|I_u\right|}
\sum_{A,B \in I_u}{\left|ModelSimRand(A, B) - GoldSimRand(A, B)\right|}
\end{align}
with $I_u$ the set of item pairs rated by user $u$ such that $\epsilon_u$ is
the average absolute error of user $u$'s similarity predictions.

@@

\section*{Conclusion}
\section*{Future Work}

\begin{thebibliography}{9}
